### 2017-01-04 introduction.tex
Motivations and Objectives revised

It is well accepted that the polar cap ionosphere is highly structured in a non-trivial manner \citep{Carlson2012}.  In addition to being of purely scientific interest, this structuring creates complications when sending radio waves through the ionosphere.  Because the structuring is highly irregular and varies on many spatial and temporal scales, unpredictable wave refraction, phase shifts, and amplitude attenuation can occur, making signals difficult to detect.  This introduces problems in any system that involves ground-to-satellite communication, such as navigation or communication.  Because of the increasing interconnectedness of modern society, any interruptions or errors in these systems can have broad impacts on infrastructure, including commercial and defense interests, as well as everyday life.

@@

The aim of this work is to examine some of the factors (such as global solar control and GDI directional dependance) behind plasma irregularity production in the F region polar cap using both observations and theory.  The majority of observational data has been found from ground based ionospheric radars, both CSR systems (Section \ref{sec:csr}) and ISR systems (Section \ref{sec:isr}), with particular emphasis on radars within the SuperDARN network, Section \ref{sec:superdarn}.  However, these datasets have also been supplemented with information from models as well as other instruments when necessary.  In addition, most experimental results are compared with predictions made by a linear fluid theory view of ionospheric plasma behavior.  Specific objectives are as follows.

The first objective of is to evaluate the extent of solar control over irregularity production in the polar F region.  This considers a global view of small-scale plasma structuring.  The occurrence of radar backscatter will be compared with both solar illumination (considered direct solar control) and IMF factors (considered indirect solar control).  This gives a macrophysical view of where and when plasma irregularities are most likely to occur.

The second objective is to investigate asymmetry in the GDI growth rate surrounding large-scale density structures.  This is a modeling approach to identifying where plasma irregularities occur surrounding a polar patch and how they can be observed by a ground-based HF radar.  The model allows certain factors to be considered that can be challenging to investigate experimentally because it is difficult to isolate particular effects.  The model used considers the direction of a large, elongated polar patch as well as its drift direction to be arbitrary in relation to the radar and independent of each other, which gives some insight as to how structuring would be expected with polar patches in orientations that are not often observed.

The third objective is to investigate the relationship between factors that control the linear GDI growth rate and small-scale plasma structuring.  This combines the theoretical consideration of where growth rate is greatest with the experimental record of where structuring is observed around a polar patch.  This is particularly important because even though plasma structuring is often observed on the trailing edge of polar patches and this is where GDI predicts the greatest growth rates, GDI theory actually predicts growth of much larger irregularities than the small scale structuring observed by HF radars.  A direct comparison of the two improves our understanding of how growth rates are related to small-scale plasma structuring.

The outline of this thesis is as follows.  Chapter \ref{sec:introduction} has provided an overview of the polar ionosphere and both macro- and microphysical plasma structuring mechanisms.  Chapter \ref{sec:paper1} will discuss plasma irregular occurrence and the extent to which solar factors control this occurrence.  Chapter \ref{sec:paper2} will consider the directional dependance of the GDI growth rate, and in particular give recommendations for the optimal orientation between a ground-based ionospheric radar and a large-scale density structure to observe small-scale structuring.  Chapter \ref{sec:paper3} will examine coincident observations made an IRS of background plasma density conditions and a CSR of small-scale plasma structuring in the polar cap and provide evidence for small-scale structuring being directionally dependent in a similar fashion to GDI predictions for large-scale plasma structuring.  Chapter \ref{sec:conclusion} will conclude this thesis and give a summary of the most significant results.  The implications of these results will be discussed along with questions that are still open and recommendations for future research.

>>>>

The plasma in the polar cap ionosphere is highly structured in a non-trivial manner \citep{Tsunoda1988,Carlson2012}.  In addition to being of purely scientific interest, this structuring creates complications when receiving radio signals through the ionosphere.  Because the structuring is highly irregular and varies on many spatial and temporal scales, unpredictable wave refraction, phase shifts, and amplitude attenuation can occur, making signals difficult to detect.  This introduces problems in any system that involves ground-to-satellite, satellite-to-ground, or over-the-horizon communication, such as navigation or communication.  Because of the increasing interconnectedness of modern society, any interruptions or errors in these systems can have broad impacts on infrastructure, including commercial and defense interests, as well as everyday life.

@@

The aim of this work is to investigate small-scale plasma irregularity production in the F-region polar cap.  In particular, factors such as direct and indirect global solar control and GDI directional dependance will be considered.  The majority of observational data has been found from ground based ionospheric radars, both CSR systems (Section \ref{sec:csr}) and ISR systems (Section \ref{sec:isr}), with particular emphasis on radars within the SuperDARN network, Section \ref{sec:superdarn}.  However, these datasets have also been supplemented with information from models as well as satellite data when necessary.  Results are investigated in the context of a linear fluid theory analysis of ionospheric plasma.

The first objective of is to evaluate the extent of solar control over irregularity production in the polar F region.  The occurrence of radar backscatter will be compared with both solar illumination (considered direct solar control) and IMF factors (considered indirect solar control).  This gives a macrophysical view of where and when plasma irregularities are most likely to occur.

The second objective is to investigate asymmetry in the GDI growth rate surrounding large-scale density structures.  A model is used to identifying where plasma irregularities occur surrounding a polar patch and how they can be observed by a ground-based HF radar.  The model considers a large, elongated polar patch drifting in an arbitrary direction relative to a radar and assumes irregularity production is directly dependent to the linear GDI growth rate.  High linear growth rates predicted by the model are compared with classic observations of stronger backscatter power trailing drifting polar patches.

The third objective is to investigate directional dependancies of small-scale plasma irregularities.  Two dimensional measurements of density gradients and electric fields are made such that the directionally dependent GDI growth rate can be calculated, which is then compared with plasma irregularity power and occurrence in a particular location.  This gives insight into whether the production of small-scale irregularities is directionally dependent, similar to the linear GDI growth rate, or if the distribution is isotropic due to other nonlinear processes.

The outline of this thesis is as follows.  Chapter \ref{sec:introduction} has provided an overview of the polar ionosphere and both macro- and microphysical plasma structuring mechanisms.  Chapter \ref{sec:paper1} presents an experimental investigation of decameter-scale plasma irregularities in the polar F-region and their solar control, with a particular focus on statistical analysis of data collected with a recently-deployed SuperDARN radar in the southern polar cap.  The same results were also presented in a recent journal article \citep{Lamarche2015}.  Chapter \ref{sec:paper2} presents a modeling study of the linear GDI growth rate surrounding a large, elongated polar patch drifting in an arbitrary direction within the FoV of a SuperDARN radar uses the results to provide insight as to optimal conditions from radars to observe irregularities.  These results have also been recently published \citep{Lamarche2016}.  Chapter \ref{sec:paper3} presents an experimental investigation of the directional dependence of small-scale plasma structures by considering simultaneous measurements of electric field and density gradients from an AMISR  system and FAIs observed by SuperDARN.  Particular emphasis is placed on if small-scale structures are directionally dependent similar to the predictions of linear GDI theory.  The same results have been submitted for publication \citep{Lamarche2017}.  Chapter \ref{sec:conclusion} presents a summary of the most significant results of this body of work, their implications for interpretation of radar observations of plasma structures in the polar cap and suggestions for future research involving coherent and incoherent scatter radar facilities in the polar regiont.
